---
title: "Bayesian analyses made easy: GLMMs in R package `brms`"
output:
  html_document:
    keep_md: yes
    code_folding: show
    df_print: paged
    highlight: tango
    number_sections: yes
    theme: united
    toc: yes
    toc_depth: 2
  pdf_document:
    toc: yes
    toc_depth: '2'
date: "April 18, 2022"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, dpi = 300)
library(tidyverse)
theme_set(theme_light(base_size = 14))
```

# Motivation

I regularly give [a course on Bayesian statistics with `R` for non-specialists](https://oliviergimenez.github.io/bayesian-stats-with-R/). To illustrate the course, we analyse data with [generalized linear, often mixed, models or GLMMs](https://bbolker.github.io/mixedmodels-misc/glmmFAQ.html). 

So far, I've been using `Jags` to fit these models. This requires some programming skills, like e.g. coding a loop, to be able to write down the model likelihood. Although students learn a lot from going through that process, it can be daunting. 

This year, I thought I'd show them the [`R` package `brms`](https://paul-buerkner.github.io/brms/) developed by [Paul-Christian Bürkner](https://paul-buerkner.github.io/). In brief, `brms` allows fitting GLMMs (but not only) in a `lme4`-like syntax within the Bayesian framework and MCMC methods with Stan.  

I'm not a Stan user, but it doesn't matter. The [vignettes](https://paul-buerkner.github.io/brms/articles/) were more than enough to get me started. I also recommend the [list of blog posts about `brms`](https://paul-buerkner.github.io/blog/brms-blogposts/). 

First things first, we load the packages we will need:
```{r}
library(brms)
library(posterior) # tools for working with posterior and prior distributions
library(R2jags) # run Jags from within R
library(lme4) # fit GLMM in frequentist framework
```

# Beta-Binomial model

The first example is about a survival experiment. We captured, marked and released 57 individuals (*total*) out of which 19 made it through the year (*alive*). 
```{r}
dat <- data.frame(alive = 19, total = 57)
dat
```

An obvious estimate of the probability of a binomial is the proportion of cases, $19/57 = 0.3333333$ here, but let's use `R` built-in functions for the sake of illustration.

## Maximum-likelihood estimation

To get an estimate of survival, we fit a logistic regression using the `glm()` function. The data are grouped (or aggregated), so we need to specify the number of alive and dead individuals as a two-column matrix on the left hand side of the model formula:
```{r}
mle <- glm(cbind(alive, total - alive) ~ 1, 
           family = binomial("logit"), # family = binomial("identity") would be more straightforward
           data = dat)
```

On the logit scale, survival is estimated at:
```{r}
coef(mle) # logit scale
```

After back-transformation using the reciprocal logit, we obtain:
```{r}
plogis(coef(mle))
```

So far, so good. 

## Bayesian analysis with `Jags`

In `Jags`, you can fit this model with:
```{r}
# model
betabin <- function(){
  alive ~ dbin(survival, total) # binomial likelihood
  logit(survival) <- beta # logit link
  beta ~ dnorm(0, 1/1.5) # prior
}
# data
datax <- list(total = 57,
              alive = 19)
# initial values
inits <- function() list(beta = rnorm(1, 0, sd = 1.5))
# parameter to monitor
params <- c("survival")
# run jags
bayes.jags <- jags(data = datax,
                      inits = inits,
                      parameters.to.save = params,
                      model.file = betabin,
                      n.chains = 2,
                      n.iter = 5000,
                      n.burnin = 1000,
                      n.thin = 1)
# display results
bayes.jags
```

## Bayesian analysis with `brms`

In `brms`, you write:
```{r}
bayes.brms <- brm(alive | trials(total) ~ 1, 
                  family = binomial("logit"), # binomial("identity") would be more straightforward
                  data = dat,
                  chains = 2, # nb of chains
                  iter = 5000, # nb of iterations, including burnin
                  warmup = 1000, # burnin
                  thin = 1) # thinning
```

You can display the results:
```{r}
bayes.brms
```

And visualize the posterior density and trace of survival (on the logit scale):
```{r}
plot(bayes.brms)
```

To get survival on the $[0,1]$ scale, we extract the MCMC values, then apply the reciprocal logit function to each of these values, and summarize its posterior distribution: 
```{r}
draws_fit <- as_draws_matrix(bayes.brms)
draws_fit
summarize_draws(plogis(draws_fit[,1]))
```

What is the prior used by default? 
```{r}
prior_summary(bayes.brms)
```

What if I want to use the same prior as in `Jags` instead? 
```{r}
nlprior <- prior(normal(0, 1.5), class = "Intercept") # new prior
bayes.brms <- brm(alive | trials(total) ~ 1, 
                  family = binomial("logit"), 
                  data = dat, 
                  prior = nlprior, # set prior by hand
                  chains = 2, # nb of chains
                  iter = 5000, # nb of iterations, including burnin
                  warmup = 1000, # burnin
                  thin = 1)
```

Double-check the prior that was used:
```{r}
prior_summary(bayes.brms)
```

# Logistic regression with covariates

In this example, we ask whether annual variation in white stork breeding success can be explained by rainfall in the wintering area. Breeding success is measured by the ratio of the number of chicks over the number of pairs. 
```{r}
nbchicks <- c(151,105,73,107,113,87,77,108,118,122,112,120,122,89,69,71,53,41,53,31,35,14,18)
nbpairs <- c(173,164,103,113,122,112,98,121,132,136,133,137,145,117,90,80,67,54,58,39,42,23,23)
rain <- c(67,52,88,61,32,36,72,43,92,32,86,28,57,55,66,26,28,96,48,90,86,78,87)
dat <- data.frame(nbchicks = nbchicks, 
                  nbpairs = nbpairs,
                  rain = (rain - mean(rain))/sd(rain)) # standardized rainfall
```

The data are grouped. 

## Maximum-likelihood estimation

You can get maximum likelihood estimates with:
```{r}
mle <- glm(cbind(nbchicks, nbpairs - nbchicks) ~ rain, 
           family = binomial("logit"), 
           data = dat)
summary(mle)
```

There is a negative effect of rainfall on breeding success:
```{r}
visreg::visreg(mle, scale = "response")
```

## Bayesian analysis with `Jags`

In Jags, we can fit the same model:
```{r}
# data 
dat <- list(N = 23, # nb of years
            nbchicks = nbchicks, 
            nbpairs = nbpairs,
            rain = (rain - mean(rain))/sd(rain)) # standardized rainfall

# define model
logistic <- function() {
  # likelihood
  for(i in 1:N){ # loop over years
    nbchicks[i] ~ dbin(p[i], nbpairs[i]) # binomial likelihood
    logit(p[i]) <- a + b.rain * rain[i] # prob success is linear fn of rainfall on logit scale
  }
  # priors
  a ~ dnorm(0,0.01) # intercept
  b.rain ~ dnorm(0,0.01) # slope for rainfall
}

# function that generates initial values 
inits <- function() list(a = rnorm(n = 1, mean = 0, sd = 5), 
                         b.rain = rnorm(n = 1, mean = 0, sd = 5))

# specify parameters that need to be estimated
parameters <- c("a", "b.rain")

# run Jags
bayes.jags <- jags(data = dat,
               inits = inits,
               parameters.to.save = parameters,
               model.file = logistic, 
               n.chains = 2,
               n.iter = 5000, # includes burn-in!
               n.burnin = 1000)

# display results
bayes.jags
```

## Bayesian analysis with `brms`

With `brms`, we write:
```{r}
bayes.brms <- brm(nbchicks | trials(nbpairs) ~ rain, 
                  family = binomial("logit"), 
                  data = dat,
                  chains = 2, # nb of chains
                  iter = 5000, # nb of iterations, including burnin
                  warmup = 1000, # burnin
                  thin = 1)
```

Display results:
```{r}
bayes.brms
```

Visualize:
```{r}
plot(bayes.brms)
```

We can also calculate the posterior probability of the rainfall effect being below zero with the `hypothesis()` function:
```{r}
hypothesis(bayes.brms, 'rain < 0')
```

These results confirm the important negative effect of rainfall on breeding success.

# Linear mixed model

In this example, we have several measurements for 33 Mediterranean plant species, specifically number of seeds and biomass. We ask whether there is a linear relationship between these two variables that would hold for all species. 

Read in data, directly from the course website:
```{r}
df <- read_csv2("https://raw.githubusercontent.com/oliviergimenez/bayesian-stats-with-R/master/slides/dat/VMG.csv")
df
```

Use relevant format for columns species Sp and biomass Vm:
```{r}
df$Sp <- as_factor(df$Sp)
df$Vm <- as.numeric(df$Vm)
```

Define numeric vector of species, for nested indexing in `Jags`:
```{r}
species <- as.numeric(df$Sp)
```

Define response variable, number of seeds:
```{r}
y <- log(df$NGrTotest)
```

Standardize explanatory variable, biomass
```{r}
x <- (df$Vm - mean(df$Vm))/sd(df$Vm)
```

Now build dataset:
```{r}
dat <- data.frame(y = y, x = x, species = species)
```

## Maximum-likelihood estimation

You can get maximum likelihood estimates for a linear regression of number of seeds on biomass with a species random effect on the intercept (partial pooling):
```{r}
mle <- lmer(y ~ x + (1 | species), data = dat)
summary(mle)
```

## Bayesian analysis with `Jags`

In `Jags`, you would use:
```{r}
# partial pooling model
partial_pooling <- function(){
  # likelihood
  for(i in 1:n){
    y[i] ~ dnorm(mu[i], tau.y)
    mu[i] <- a[species[i]] + b * x[i]
  }
  for (j in 1:nbspecies){
    a[j] ~ dnorm(mu.a, tau.a)
  }
  # priors
  tau.y <- 1 / (sigma.y * sigma.y)
  sigma.y ~ dunif(0,100)
  mu.a ~ dnorm(0,0.01)
  tau.a <- 1 / (sigma.a * sigma.a)
  sigma.a ~ dunif(0,100)
  b ~ dnorm(0,0.01)
}

# data
mydata <- list(y = y, 
               x = x, 
               n = length(y),
               species = species,
               nbspecies = length(levels(df$Sp)))

# initial values
inits <- function() list(mu.a = rnorm(1,0,5), 
                         sigma.a = runif(0,0,10), 
                         b = rnorm(1,0,5), 
                         sigma.y = runif(0,0,10))

# parameters to monitor
params <- c("mu.a", "sigma.a", "b", "sigma.y")

# call jags to fit model
bayes.jags <- jags(data = mydata,
                   inits = inits,
                   parameters.to.save = params,
                   model.file = partial_pooling,
                   n.chains = 2,
                   n.iter = 5000,
                   n.burnin = 1000,
                   n.thin = 1)
bayes.jags
```

## Bayesian analysis with `brms`

In `brms`, you write:
```{r}
bayes.brms <- brm(y ~ x + (1 | species), 
                  data = dat,
                  chains = 2, # nb of chains
                  iter = 5000, # nb of iterations, including burnin
                  warmup = 1000, # burnin
                  thin = 1)
```

Display the results:
```{r}
bayes.brms
```

You can extract a block of fixed effects:
```{r}
summary(bayes.brms)$fixed
```

And a block of random effects:
```{r}
summary(bayes.brms)$random
```

And visualize:
```{r}
plot(bayes.brms)
```

# GLMM with Poisson distribution

This example is from Jason Matthiopoulos' excellent book [*How to be a quantitative ecologist*](http://greenmaths.st-andrews.ac.uk/).
A survey of a coral reef uses 10 predefined linear transects covered by divers once every week. The response variable of interest is the abundance of a particular species of anemone as a function of water temperature. Counts of anemones are recorded at 20 regular line segments along the transect.

Let's simulate some data:
```{r}
set.seed(666)
transects <- 10
data <- NULL
for (tr in 1:transects){
  # random effect (intercept)
  ref <- rnorm(1,0,.5) 
  # water temperature gradient
  t <- runif(1, 18,22) + runif(1,-.2,0.2)*1:20 
  # Anemone gradient (expected response)
  ans <- exp(ref -14 + 1.8 * t - 0.045 * t^2) 
  # actual counts on 20 segments of the current transect
  an <- rpois(20, ans) 
  data <- rbind(data, cbind(rep(tr, 20), t, an))
}

data <- data.frame(Transect = data[,1],
                   Temperature = data[,2],
                   Anemones = data[,3])
```

Standardize temperature:
```{r}
boo <- data$Temperature
data$Temp <- (boo - mean(boo)) / sd(boo)
head(data)
```

## Maximum-likelihood estimation

With `lme4`, you write:
```{r}
fit_lme4 <- glmer(Anemones ~ Temp + I(Temp^2) + (1 | Transect), 
                  data = data, 
                  family = poisson)
summary(fit_lme4)
```

## Bayesian analysis with `Jags`

In `Jags`, we fit the corresponding GLMM with:
```{r}
model <- function() {
  for (i in 1:n){
    count[i] ~ dpois(lambda[i])
    log(lambda[i]) <- a[transect[i]] + b[1] * x[i] + b[2] * pow(x[i],2)
  }
  for (j in 1:nbtransects){
    a[j] ~ dnorm (mu.a, tau.a)
  }
  mu.a ~ dnorm (0, 0.01)
  tau.a <- pow(sigma.a, -2)
  sigma.a ~ dunif (0, 100)
  b[1] ~ dnorm (0, 0.01)
  b[2] ~ dnorm (0, 0.01)
}

dat <- list(n = nrow(data), 
            nbtransects = transects, 
            x = data$Temp, 
            count = data$Anemones, 
            transect = data$Transect)

inits <- function() list(a = rnorm(transects), 
                         b = rnorm(2), 
                         mu.a = rnorm(1), 
                         sigma.a = runif(1))

par <- c ("a", "b", "mu.a", "sigma.a")

fit <- jags(data = dat, 
            inits = inits, 
            parameters.to.save = par, 
            model.file = model,
            n.chains = 2, 
            n.iter = 5000, 
            n.burn = 1000)

round(fit$BUGSoutput$summary[, -c(4,6)], 3)
```

## Bayesian analysis with `brms`

What about with `brms`?
```{r}
bayes.brms <- brm(Anemones ~ Temp + I(Temp^2) + (1 | Transect), 
                  data = data,
                  family = poisson("log"),
                  chains = 2, # nb of chains
                  iter = 5000, # nb of iterations, including burnin
                  warmup = 1000, # burnin
                  thin = 1)
```

Display results:
```{r}
bayes.brms
```

Visualize:
```{r}
plot(bayes.brms)
```

We can assess the quality of fit of this model:
```{r}
pp_check(bayes.brms, ndraws = 100, type = 'ecdf_overlay')
```

As expected, the fit is almost perfect because we simulated data from this very model. 

What if we'd like to test the effect of temperature using WAIC? 

We fit a model with no effect of temperature:
```{r}
bayes.brms2 <- brm(Anemones ~ 1 + (1 | Transect), 
                  data = data,
                  family = poisson("log"),
                  chains = 2, # nb of chains
                  iter = 5000, # nb of iterations, including burnin
                  warmup = 1000, # burnin
                  thin = 1)
```

Then we compare both models, by ranking them with their WAIC or using ELPD:
```{r}
(waic1 <- waic(bayes.brms)) # waic model w/ tempterature
(waic2 <- waic(bayes.brms2)) # waic model wo/ tempterature
loo_compare(waic1, waic2)
```

## Conclusions

In all examples, you can check that i) there is little or no difference between `Jags` and `brms` numeric summaries of posterior distributions, and ii) maximum likelihood parameter estimates (`glm()`, `lmer()` and `glmer()` functions) and posterior means/medians are very close to each other. 

In contrast to Jags, Nimble or Stan, you do not need to code the likelihood yourself in `brms`, as long as it is available in the package. You can have a look to a list of distributions with `?brms::brmsfamily`, and there is also a possibility to define custom distributions with `custom_family()`. 

Note I haven't covered diagnostics of convergence, more [here](https://www.rensvandeschoot.com/brms-wambs/). You can also integrate `brms` outputs in a [tidy workflow with `tidybayes`](http://mjskay.github.io/tidybayes/articles/tidy-brms.html), and use the `ggplot()` magic.

Now I guess it's up to the students to pick their favorite Bayesian tool.  
